---
title: "Analysis of fold changes of CRISPR-targeted genes from the ENCODE data set"
author: "Frederik Ziebell"
date: "29 3 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
options(width = 150)
```

# preparations
```{r}
suppressPackageStartupMessages({
  library("AnnotationHub")
  library("ensembldb")
  library("glue")
  library("cowplot")
  library("tidyverse")
})
theme_set(theme_cowplot())

# Annotate ENSEMBL genes with gene symbols
ah <- AnnotationHub()
edb <- query(ah,c("EnsDb","sapiens",90))[[1]]
gene_symbols <- genes(edb, return.type="data.frame") %>%
  remove_rownames() %>%
  select(gene_id, hgnc_symbol=gene_name)
```

# download data
```{r}
# all files related to CRISPR screens
files <- read_lines("files.txt")

# metadata for all files that can be downloaded
files_metadata <- read_tsv("metadata.tsv", guess_max=0) %>%
  `colnames<-`(str_to_lower(str_replace_all(colnames(.)," ","_")))

# filter for expression and differential testing results of gene deletions
metadata_filtered <- files_metadata %>%
  filter(file_format=="tsv") %>%
  filter(output_type=="differential expression quantifications") %>%
  filter(biosample_genetic_modifications_categories=="deletion") %>%
  mutate(targets=str_replace(str_replace(str_replace(biosample_genetic_modifications_targets,"/targets/|",""),"-human/",""),"Non-specific%20target%20control","control")) %>%
  select(file_accession, output_type, experiment_accession,assay,biosample_term_name,targets) %>%
  arrange(targets,output_type)
  
# compute files to be downloaded
# for download, use command: 'xargs -L 1 curl -O -L < download_files.txt'
# all files have already been downloaded in the folder 'data'
str_subset(files,str_c(metadata_filtered$file_accession,collapse="|")) %>%
  write_lines("download_files.txt")
```

# load differential expression results
```{r}
# metadata of differential expression files
metadata_de <- metadata_filtered %>%
  filter(str_detect(output_type,"expression"))
head(metadata_de)

# load differential testing results files
results_de <- list.files("data",pattern=".tsv$") %>%
  str_subset(str_c(metadata_de$file_accession,collapse="|")) %>%
  map(function(f) {read.delim(glue("data/{f}")) %>%
      mutate(filename=f)})

# extract testing results from both types of results files
results_de_type1 <- which(map_int(results_de,ncol) %in% c(9,10)) %>%
  results_de[.] %>%
  reduce(bind_rows)
head(results_de_type1)

results_de_type2 <- which(map_int(results_de,ncol)==15) %>%
  results_de[.] %>%
  reduce(bind_rows)
head(results_de_type2)

# get residual RNA levels for CRISPR-targeted genes
residual_rna_levels <- bind_rows(
  # results of type 1
  results_de_type1 %>%
    mutate(base_file=str_replace(filename,".tsv","")) %>%
    left_join(metadata_de, by=c("base_file"="file_accession")) %>%
    mutate(gene_id=str_replace(id,"\\.[:number:]+","")) %>%
    left_join(gene_symbols, by="gene_id") %>%
    filter(hgnc_symbol==targets) %>%
    select(target_name=hgnc_symbol,experiment_id=experiment_accession, parent_line=biosample_term_name, fold_change=foldChange),
  # results of type 2
  results_de_type2 %>%
    mutate(base_file=str_replace(filename,".tsv","")) %>%
    left_join(metadata_de, by=c("base_file"="file_accession")) %>%
    filter(gene==targets) %>%
    mutate(fold_change=2^log2.fold_change.) %>%
    select(target_name=gene,experiment_id=experiment_accession, parent_line=biosample_term_name, fold_change)
) 
```

# inspect results
```{r}
residual_rna_levels %>%
  ggplot(aes(target_name, fold_change)) +
  geom_point(alpha=.6) +
  geom_abline(slope=0, intercept=1, linetype="33") +
  labs(x="targeted gene", y="fold-change vs control") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))
```

This looks odd. Almost all CRISPR-targeted genes show *increased* RNA levels. Let's re-compute fold changes from the expression levels and compare with reported fold-changes.

```{r}
results_de_type2 <- results_de_type2 %>%
  mutate(base_file=str_replace(filename,".tsv","")) %>%
  left_join(metadata_de, by=c("base_file"="file_accession")) %>%
  filter(gene==targets) %>%
  select(gene, sample_1, sample_2, value_1, value_2, log2fc_reported=log2.fold_change.) %>%
  mutate(log2fc_recomputed=log2(value_1/value_2))

head(results_de_type2)

results_de_type2 %>%
  ggplot(aes(log2fc_recomputed,log2fc_reported)) +
  geom_abline(intercept=0, slope=-1, color="red") +
  geom_point(alpha=.6) +
  xlim(-3,3) +
  ylim(-3,3) +
  coord_fixed() +
  geom_hline(yintercept=0, linetype="33")
```

Since "sample_2" is annotated as control, it tells us that value_2 has to be in the denominator. However, the reported log2-fold-change is exactly the opposite and almost all reported log2-fold-changes are positive, indicating that the wrong fold change has been computed.

For the other results type, we see a similar pattern, suggesting that fold changes are incorrect as well.
```{r}
results_de_type1 <- results_de_type1 %>%
  mutate(base_file=str_replace(filename,".tsv","")) %>%
  left_join(metadata_de, by=c("base_file"="file_accession")) %>%
  mutate(gene_id=str_replace(id,"\\.[:number:]+","")) %>%
  left_join(gene_symbols, by="gene_id") %>%
  filter(hgnc_symbol==targets) %>%
  select(gene_id=id, mean_a=baseMeanA, mean_b=baseMeanB, log2fc_reported=log2FoldChange) %>%
  mutate(log2fc_recomputed=log2(mean_a/mean_b)) 

results_de_type1 %>%
  ggplot(aes(log2fc_recomputed,log2fc_reported)) +
  geom_abline(intercept=0, slope=-1, color="red") +
  geom_point(alpha=.6) +
  xlim(-3,3) +
  ylim(-3,3) +
  coord_fixed() +
  geom_hline(yintercept=0, linetype="33")
```

# corrected fold-changes
```{r}
residual_rna_levels %>%
  ggplot(aes(target_name, 1/fold_change)) +
  geom_point(alpha=.6) +
  geom_abline(slope=0, intercept=1, linetype="33") +
  labs(x="targeted gene", y="fold-change vs control") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))
```

Almost all cell lines show reduced RNA-levels. This seems more reasonable.  
